\section{Layer~7 – Emergent Self‑Awareness}
\label{sec:layer7}

% ----- SUPERKRACHT -----
\begin{tcolorbox}[colback=blue!5, colframe=blue!60, title=\textbf{Superpower: The Watcher}]
\begin{center}
\begin{tikzpicture}
  \node[draw, circle, fill=green!10, minimum size=1.5cm, font=\bfseries\Huge] {W};
\end{tikzpicture}
\textit{“How am I doing right now?”}
\end{center}
\end{tcolorbox}

Pixel is slowly growing up.  He has all those layers, all those
connections, but no idea whether the whole system is healthy.  Layer~7
builds a \textbf{dashboard} – an internal overview that shows at a
glance whether all the parts are still working well together.

\begin{tcolorbox}[colback=yellow!5, colframe=yellow!80, title=Real‑life analogy]
A modern aeroplane has hundreds of sensors measuring temperature,
pressure, speed, and engine status.  The pilot sees on a screen
whether everything is green, or whether a red light is flashing
somewhere.  Pixel’s dashboard does exactly the same, but for his own
“thought processes”.
\end{tcolorbox}

\textbf{Pixel’s inner monologue.}
If a functional cluster suddenly becomes unreliable, Pixel notices it
by a falling “health indicator”.  He can then gently correct course
before real errors happen.  This is not consciousness as we know it –
Pixel feels no joy or pain – but it is a primitive form of
\textbf{self‑awareness}: the system knows how it is doing.

Self‑awareness is an important safety mechanism.  An AI that can
recognise its own confusion can raise a red flag before it makes
dangerous decisions.