\section{Python verification}\label{sec:verification}

We implemented the functor \(\mathbf{D}\) in Python using the reference APEIRON Layer~1.
The core logic of computing \(\mathcal{L}(o)\) and extracting the induced subgraph is shown in Listing~\ref{lst:functor}.

\begin{lstlisting}[language=Python, caption={Core of the functor implementation}, label={lst:functor}]
def downstream_reachable(fwd_edges, root_id):
    """Compute L(o): all nodes reachable from root_id."""
    visited = set()
    stack = [root_id]
    while stack:
        cur = stack.pop()
        if cur in visited:
            continue
        visited.add(cur)
        for nxt in fwd_edges.get(cur, []):
            if nxt not in visited:
                stack.append(nxt)
    return visited

def build_functor_graph(seeds):
    """Build D(o) for each seed observable."""
    objects, fwd_edges = build_decomposition_graph(seeds)
    result = {}
    for obs in seeds:
        lset = downstream_reachable(fwd_edges, obs.id)
        edges = [(s,t) for s in lset for t in fwd_edges.get(s,[]) if t in lset]
        result[obs.id] = (lset, edges)
    return result
\end{lstlisting}

The complete script \texttt{functorial\_emergence\_verify.py} is available in the supplementary material and can be run on any machine with the APEIRON package installed.

\subsection{Test setup}
To thoroughly validate the functor, we employ a diverse set of twelve observables, comprising:
\begin{itemize}
\item Atomic integers: \(1\), \(2\), \(3\), and the prime \(7\).
\item Composite sets: \(\{1,2\}\), \(\{1,2,3\}\), and \(\{1,2,3,4\}\).
\item Edge cases: the empty set, an empty string, a repetitive string (\texttt{"a"*200}), and a random 200‑character string.
\item A list \texttt{[1,2]} as an example of a non‑atomic sequence.
\end{itemize}
All observables are instantiated via the UltimateObservable class and their atomicity is computed using the four primary frameworks (Boolean, Measure, Categorical, Information).
The decomposition graph is built downstream, and for each observable \(o\) we collect the set \(\mathcal{L}(o)\) and count the internal edges.

\subsection{Interpretation}
\Cref{tab:functor} presents the results.
Every atomic observable, regardless of its type (integer, empty set, string), yields a singleton graph (\(\lvert\mathcal{L}(o)\rvert = 1\), zero edges).
In contrast, every composite object produces a non‑trivial graph whose size grows with the complexity of the object.
For example, the set \(\{1,2\}\) decomposes into \(\{1\}\) and \(\{2\}\), giving a graph with three vertices and two edges; the larger set \(\{1,2,3,4\}\) generates a graph with 15 vertices and 50 edges.
This pattern confirms Lemma~\ref{lem:separated_map} over a wide range of data, and the fully automated, machine‑generated table guarantees reproducibility.

\begin{table}[htbp]
\centering
\caption{Functor verification. An observable is \emph{Separated} if it is multi‑axially atomic. \(\vert\mathcal{D}(o)\vert\) is the number of vertices in \(\mathbf{D}(o)\).}
\label{tab:functor}
\input{tables/decomposition_table.tex}
\end{table}