\begin{abstract}
As artificial intelligence systems grow in complexity and autonomy, the need for
machine‑checkable guarantees about their foundational properties becomes
paramount.  This paper presents a multi‑prover framework that formally verifies
the \emph{atomicity}—the irreducibility—of the elementary representational
units within the \Apeiron{} Framework, a seventeen‑layer architecture for
non‑anthropocentric knowledge genesis.  We integrate four theorem provers
(SymPy, Z3, Lean~4, and Coq) into a unified pipeline, \texttt{self\_proving.py},
that automatically generates proof obligations in each prover's native language,
executes them, and exports machine‑checkable JSON certificates.

We evaluate the system on a benchmark of 47 observables comprising prime
numbers, composite numbers, finite sets, and repetitive strings.  A rigorous
verification of each prover's output reveals that \textbf{Z3} is the only
fully automatic, sound prover for Boolean atomicity on integers and sets,
correctly distinguishing atoms from non‑atoms across all test cases.
\textbf{SymPy} serves as a fast, lightweight consistency check but is not
sound for atomicity.  \textbf{Lean~4} and \textbf{Coq} generate correct
proofs for integers via computational reflection (\texttt{native\_decide} /
\texttt{vm\_compute}), but their backends for non‑integer observables are
currently placeholders and must not be trusted.  We define a \emph{trust
hierarchy} that assigns formal weight to each prover accordingly, and we
generate certificates that reflect this hierarchy.  Our work demonstrates a
pragmatic, verifiable multi‑prover architecture and openly discusses the
limitations that remain, providing a realistic foundation for formal methods
in autonomous AI systems.

\textbf{Keywords:} Formal verification, theorem proving, SMT, interactive
proof assistants, AI alignment, epistemic foundations, atomicity,
\Apeiron{} Framework.
\end{abstract}