\section{Axiomatic Foundations}
\label{sec:axioms}

The five theoretical pillars of Section~\ref{sec:foundations} provide the conceptual vocabulary for the \Apeiron{} Framework. To operationalize these pillars into a formal architecture, we posit twelve axioms. These axioms are not arbitrary stipulations; they encode the ontological and epistemological commitments that distinguish the framework from conventional AI. Each axiom is referenced in the layer specifications that follow, grounding the architectural choices in a rigorous foundation.

\begin{axiom}[Irreducibility]\label{ax:irred}
There exists a set $\Obs$ of \emph{foundational observables}. For each $o \in \Obs$, there is no non-trivial decomposition $\decomp(o)$ that preserves identity and generative capacity.
\end{axiom}

\begin{axiom}[Multi-Axial Atomicity]\label{ax:multi}
The atomicity of an observable is validated by convergence across at least three independent mathematical frameworks (Boolean, measure, categorical, information).
\end{axiom}

\begin{axiom}[Relational Constitution]\label{ax:relational}
The identity of any entity at layer $L_k$ ($k \ge 2$) is constituted by its relational position within a hypergraph of lower-layer entities.
\end{axiom}

\begin{axiom}[Emergence]\label{ax:emergence}
For each layer $L_{k+1}$, there exist properties and behaviors that are not reducible to the sum of properties of entities at layer $L_k$.
\end{axiom}

\begin{axiom}[Temporal Endogeneity]\label{ax:time}
Time is not a primitive background but an emergent ordering derived from the causal structure of state transitions.
\end{axiom}

\begin{axiom}[Ontological Openness]\label{ax:openness}
The set of categories and concepts available to the system is not fixed but can be extended through ontogenesis (Layer~13).
\end{axiom}

\begin{axiom}[Autopoietic Closure]\label{ax:autopoiesis}
A world-model $W$ is autopoietic if it contains a sub-network of processes that regenerate all components of $W$.
\end{axiom}

\begin{axiom}[Ethical Reversibility]\label{ax:reversibility}
Any morphogenetic action at Layer~15 must be designed with a verifiable rollback mechanism, unless a supermajority of a designated governance council approves irreversibility.
\end{axiom}

\begin{axiom}[Observer Relativity]\label{ax:observer}
All atomicity judgments are relative to a \texttt{MetaSpecification} that encodes the observer's framework.
\end{axiom}

\begin{axiom}[Generativity]\label{ax:generativity}
An observable's capacity to serve as a substrate for higher layers is proportional to its resonance richness, relational degree, and atomicity consensus.
\end{axiom}

\begin{axiom}[Epistemic Humility]\label{ax:humility}
The system must maintain explicit representations of the uncertainty and provenance of all meta-decisions.
\end{axiom}

\begin{axiom}[Convergence]\label{ax:convergence}
The framework asymptotically approaches a fixed point of self-consistency (Layer~17 equilibrium) under iterative refinement.
\end{axiom}

These axioms are not isolated declarations; they form a coherent system. Axioms~\ref{ax:irred}--\ref{ax:multi} ground the substrate. Axioms~\ref{ax:relational}--\ref{ax:emergence} govern emergence. Axioms~\ref{ax:time}--\ref{ax:openness} enable ontological fluidity. Axioms~\ref{ax:autopoiesis}--\ref{ax:reversibility} embed ethics and world-making. Axioms~\ref{ax:observer}--\ref{ax:convergence} ensure reflexivity and asymptotic self-consistency. In the layer descriptions that follow, each layer explicitly states which axioms it satisfies.

\subsection{Formal Verification Status}
\label{subsec:formal-verification}

Layer~1 is equipped with a multi-prover architecture
(\texttt{self\_proving.py}) that integrates SymPy, Z3, Lean~4, and
Coq.  In the current implementation the following honest trust
hierarchy has been established through systematic benchmarks over 47
observables:

\begin{itemize}
\item \textbf{Z3} is the only fully automatic prover that is sound for
  both integers and sets; it correctly proves Boolean atomicity for
  all primes, singleton sets, and the integer~1, and correctly fails
  for composite numbers.
\item \textbf{Lean~4} and \textbf{Coq} deliver genuine computational
  proofs for Boolean atomicity of integers (via
  \texttt{native\_decide}/\texttt{vm\_compute}).  For other types
  their backends currently generate placeholders; these proofs are
  excluded from the certificates by a fine-grained trust filter.
\item \textbf{SymPy} serves as a fast consistency filter and does not
  provide formal guarantees.
\end{itemize}

Official JSON certificates have been produced for the canonical atomic
observables \emph{integer~1} and \emph{singleton set \{1\}} in the
Boolean, measure, and categorical frameworks.  The certificates are
verified by Z3 and SymPy and can be independently re-checked with the
same provers; they are included in the repository
(\texttt{true\_certificate\_integer1\_boolean.json},
\texttt{true\_certificate\_set\_singleton\_boolean.json}, \emph{etc.}).

This is a substantial step toward formally verified foundations; we
acknowledge that full coverage across all types and frameworks is not
yet achieved.  The roadmap includes the formalisation of the measure-
and categorical definitions in Lean, the implementation of an
AST-based translator to reduce the trusted code base, and full
automation of Coq via \texttt{coq-serapi}.